Secure Composition of Robust and Optimising Compilers

### 2023-07-10 popl/main.tex
intro.

\gls{ms} is a security property of programs that, e.g., ensures that there are no out-of-bounds accesses as well as no reads or writes to uninitialized memory.
Without memory safety, private data can easily leak or be corrupted~\cite{lemay2021ccc}.
Because of this, \gls{ms} is a requirement for programs to attain an even stronger security property, namely \gls{cct}.
Programs that attain \gls{cct} are secure with respect to timing attacks, and this is the gold standard for secure applications, including cryptographic ones since timing attacks can expose secrets via a simple performance analysis~\cite{kocher1996timing}.
Timing attacks have become more and more relevant recently, especially given the era of cloud-computing~\cite{aviram2010cloudtime,kumar2019cloudsecsurvey}, where programs share the same hardware but are supposed to run in a sandboxed environment that does not leak secrets.
Unfortunately, several attacks within such a cloud environment have been demonstrated as viable strategies to extract secrets~\cite{mehmet2015getoff,flowers2022zeroday,atya2019catchme,venkatanathan2015placevul}.

Even though mitigations for such attacks exist~\cite{bond2017vale,almeida2017jasmin}, the literature shows that some compilers may optimize these mitigations away~\cite{barthe2018sec}.
This is of course not what programmers expect and necessitates the need for a {\em secure} compiler that ensures such mitigations persist in the code they generate.
More generally, secure compilers should preserve security properties even when the program runs in a hostile context such as the cloud one, i.e., where it interacts with other possibly malicious programs.
An example of such secure compilers are those that ascribe to the theory of robust compilation~\cite{abate2019jour,abate2021extacc,patrignani2021rsc}, which takes this interaction between components and potentially malicious contexts into account.

Unfortunately, it is unclear if and how robust compilation supports compositionality, and compilation passes are typically developed as a standalone module to be plugged into a compilation pipeline and composed with other passes.
Different kinds of compilation passes exist, e.g., there are usually several compilation passes that perform optimisations~\cite{lattner2004llvm,googlev8,androidstudio,kuepper2023cryptopt,manjikian1997fusion,wegman1991ccp} or source-code instrumentations~\cite{nagarakatte2009soft,nagarakatte2010cets,akritidis2009baggy,dhumbumroong2020boundwarden,jung2021pico,nam2019framer,shankaranarayana2023tailcheck,younan2010paricheck,zhou2023fatptrs,bond2017vale,almeida2017jasmin,kuepper2023cryptopt,cauligi2019fact} that enforce that the compiled program fulfils a property of interest.
Without a framework for composing secure compiler passes, it is not possible to enable separation of concerns, e.g., to have a secure compilation pass that ensures \gls{ms} that is developed independently of another secure pass for \gls{cct} as showcased in \Cref{ex:strncpy}.
\begin{example}[strncpy]\label{ex:strncpy}
Consider the \texttt{C} program below that copies a null-terminated string \texttt{src} into \texttt{dst} up to a length of \texttt{n}:

@@

For sake of simplicity, assume that the length of the array \texttt{dst} points to is at least as large as \texttt{n} as well as \texttt{src} terminated by \texttt{'\textbackslash 0'}.
Even under these assumptions, there is a subtle \gls{ms} bug in the program: The bounds check \texttt{i < n} should happen {\it prior} to the access to memory location \texttt{x[i]}:
If \texttt{n} is at least as large as the length of the array \texttt{src}, the memory location past the last element will be read.
A compilation pass that enforces spatial memory safety, such as Softbounds~\cite{nagarakatte2009soft} or BaggyBounds~\cite{akritidis2009baggy}, could have prevented this accidental error.

Unfortunately, fixing \gls{ms} is not enough to have a secure program, because \texttt{strncpy} is susceptible to timing attacks.
This is because the loop can terminate early: As soon as the string-terminating character \texttt{'\textbackslash 0'} is encountered, the loop exits.
Thus, the program execution time is proportional to the length of the array \texttt{src} points to.
Often, programmers rewrite programs like these to have loops that run with a fixed number of iterations.
However, compiler passes are capable of rewriting such programs~\cite{cauligi2019fact} and a modification by hand is not necessary.

These fixes alone are not enough to make this program secure in a robust setting, i.e., a malicious context could trigger a violation of either \gls{ms} or \gls{cct}.
Consider a program context \texttt{strncpy(1729,dst,src)} that calls \texttt{strncpy} with pointers \texttt{dst} and \texttt{src}, where \texttt{src} is some uninitialized memory.
This leads to a series of reads from uninitialised memory, which is an immediate \gls{ms} issue that can have devastating real-world implications~\cite{uninit-0,uninit-1,uninit-2,uninit-3,uninit-4}.
An easy fix to sidestep this is to use non-nullable pointers as parameters instead, e.g., as presented in Cyclone~\cite{jim2002cyclone}.
With this, a secure compiler needs to ensure that no malicious target-level context can supply unitialised memory and thereby sidestep the guarantees of the source language, e.g., by inserting appropriate checks prior to entering the body of \texttt{strncpy}.
\end{example}

This paper introduces a framework for reasoning about the composition of secure and optimising compiler passes and it showcases the power of this framework by instantiating it on a multi-pass compilation chain.
To this end, this paper first discusses how to compose security properties, such as temporal and spatial memory safety into general memory safety, and cryptographic constant time.
Then, this paper defines several secure compiler passes, where each is either preserving a different security property (e.g., temporal or spatial memory safety) or performing a security-preserving optimisation, (e.g., applying constant-folding or dead-code elimination).
Finally, this paper shows how to compose these secure compiler passes into a multi-pass compilation chain which is proven to provide end-to-end preservation of general memory safety and cryptographic constant time.
Crucially, this end-to-end security preservation is obtained by having each compilation pass (i) preserve a sub-part of the overall security property individually and then (ii) compose each preserved security property as dictated by the framework of this paper.
The results showcase how the framework allows the kind of formal security reasoning that compiler writers already want (and do), obtaining precise, compositional security reasoning while providing minimal (and modular) proof effort.

@@

  \item %
        %\Cref{sec:compcomp} studies different forms of secure compiler compositions.
        This paper takes the secure compilation framework of~\citep{abate2019jour} and extends it to reason about the security of all different known forms of compiler composition (\Cref{sec:compcomp}).
        %Concretely, it investigates upper and lower compositions, as well as the sequential composition.
        For this, this paper studies compilers with multiple inputs or multiple outputs as well as the sequential composition of compilers, as typically done when engineering compilation chains.
        %
        This paper proves that starting from two compilers that preserve two properties, their composition preserves the intersection of those properties.
        For example, a compiler that is secure with respect to \gls{ms} can be composed with a compiler that is secure with respect to \gls{cct}, resulting in a compiler that is secure with respect to the composition of \gls{ms} and \gls{cct}.
%       \MPin{
%      		find way to connect to previous contrib
%       }
        % 
        Finally, this paper provides an important corollary that demonstrates that the order of composition of sequential compiler passes is irrelevant for the resulting security.
        % 
        This corollary is crucial for reordering compiler optimisation passes as well as secure compilation passes and thus generating secure and efficient code.

  \item %
        This paper presents a case-study showcasing the conjunction of the previous contributions (\Cref{sec:casestud:defs,sec:casestud:rtp}).
        To this end, it presents a compilation chain consisting of several passes.
        The chain ultimately preserves a combination of \gls{ms} and \gls{cct} by means of composing the individual, secure passes concerning \gls{tms}, \gls{sms}, and a stricter variant of \gls{cct}, respectively.
        Furthermore, the chain includes two passes for \gls{ms} that are optimising: One performs a simple \gls{dce} and the other \gls{cf}.
        The formalisation of this case study showcases the power of the presented framework.
        Namely, the divide-and-conquer approach to software engineering is a viable strategy even for the developement of secure compilers.
%        This demonstrates that the usage of the presented framework is both simpler and more expressive compared to one monolithic security proof of the whole compilation chain.

>>>>

\gls{ms} is a security property that ensures that there are no out-of-bounds accesses as well as no reads or writes to uninitialized memory.
Programs that do not attain memory safety, can easily have their private data be leaked to attackers or be corrupted~\cite{lemay2021ccc}.
Because of this, \gls{ms} is a requirement for programs to attain an even stronger security property, namely \gls{cct}.
Programs that attain \gls{cct} are secure with respect to timing attacks, and this is the gold standard for secure applications, including cryptographic ones, since timing attacks can expose secrets via a simple performance analysis~\cite{kocher1996timing}.
Timing attacks have become more and more relevant recently, especially given the era of cloud-computing~\cite{aviram2010cloudtime,kumar2019cloudsecsurvey}, where programs share the same hardware but are supposed to run in a sandboxed environment that does not leak secrets.
Unfortunately, several attacks within such a cloud environment have been demonstrated as viable strategies to extract secrets~\cite{mehmet2015getoff,flowers2022zeroday,atya2019catchme,venkatanathan2015placevul}.

Even though mitigations for such attacks exist~\cite{bond2017vale,almeida2017jasmin}, some compilers may optimize these mitigations away~\cite{barthe2018sec}.
This is of course not what programmers expect and to ensure such mitigations persist in the compiled code, one can use a {\em secure} compilers.
More generally, secure compilers should preserve security properties even when the program runs in a hostile context such as the cloud one, i.e., where it interacts with other possibly malicious programs.
An example of such secure compilers are those that ascribe to the theory of robust compilation~\cite{abate2019jour,abate2021extacc,patrignani2021rsc}, which takes this interaction between components and potentially malicious contexts into account.

Unfortunately, it is unclear if and how robust compilation supports compositionality, and compilation passes are typically developed as a standalone module to be plugged into a compilation pipeline and composed with other passes.
Different kinds of compilation passes exist, e.g., there are usually several compilation passes that perform optimisations~\cite{lattner2004llvm,googlev8,androidstudio,kuepper2023cryptopt,manjikian1997fusion,wegman1991ccp} or source-code instrumentations~\cite{nagarakatte2009soft,nagarakatte2010cets,akritidis2009baggy,dhumbumroong2020boundwarden,jung2021pico,nam2019framer,shankaranarayana2023tailcheck,younan2010paricheck,zhou2023fatptrs,bond2017vale,almeida2017jasmin,kuepper2023cryptopt,cauligi2019fact} that enforce that the compiled program fulfils a property of interest.
Without a framework for composing secure compiler passes, it is not possible to enable separation of concerns, e.g., to have a secure compilation pass that ensures \gls{ms} that is developed independently of another secure pass for \gls{cct} as suggested before and as showcased in \Cref{ex:strncpy}.
\begin{example}[strncpy]\label{ex:strncpy}
Consider the \texttt{C} program below that copies a null-terminated string \texttt{src} into \texttt{dst} up to a length of \texttt{n}:

@@

For sake of simplicity, assume that the length of the array \texttt{dst} points to is at least \texttt{n}, and \texttt{src} is terminated by \texttt{'\textbackslash 0'}.
Even under these assumptions, there is a subtle \gls{ms} vulnerability in the program: The bounds check \texttt{i < n} should happen {\it before} the access to memory location \texttt{x[i]}:
If \texttt{n} is at least as large as the length of the array \texttt{src}, the memory location past the last element will be read.
To prevent this vulnerability, one can use a compilation pass that enforces spatial memory safety, such as Softbounds~\cite{nagarakatte2009soft} or BaggyBounds~\cite{akritidis2009baggy}.

Unfortunately, fixing \gls{ms} is not enough to have a secure program, because \texttt{strncpy} is susceptible to timing attacks.
This is because the loop can terminate early: As soon as the string-terminating character \texttt{'\textbackslash 0'} is encountered, the loop exits.
Thus, the program execution time is proportional to the length of the array that \texttt{src} points to.
% Often, programmers rewrite programs like these to have loops that run with a fixed number of iterations.
Also in this case there exist compiler passes that can rewrite such programs~\cite{cauligi2019fact} so that they are no longer vulnerable to timing attacks.

These fixes alone are not enough to make this program secure in a robust setting, i.e., a malicious context could trigger a violation of either \gls{ms} or \gls{cct}.
Consider a program context that calls \texttt{strncpy} with an argument for \texttt{src} that points to uninitialized memory.
This leads to a series of reads from uninitialised memory, which is an immediate \gls{ms} issue that can have devastating real-world implications~\cite{uninit-0,uninit-1,uninit-2,uninit-3,uninit-4}.
An easy fix to sidestep this is to use non-nullable pointers as parameters instead, e.g., as presented in Cyclone~\cite{jim2002cyclone}.
With this, a secure compiler needs to ensure that no malicious target-level context can supply unitialised memory and thereby sidestep the guarantees of the source language, e.g., by inserting appropriate checks prior to entering the body of \texttt{strncpy}.
\MPin{
	do we have examples of such wrapper-chesk inserting secure compilers?
	mswasm compiler ... not really.
	or we could say that another solution is  using capabilities, and then the mswasm compiler can be used
}
\end{example}

This paper introduces a framework for reasoning about the composition of secure and optimising compiler passes akin to those of \Cref{ex:strncpy} and it showcases the power of this framework by instantiating it on a multi-pass compilation chain.
To this end, this paper first discusses how to compose security properties, such as temporal and spatial memory safety into general memory safety, and cryptographic constant time.
\MP{
	acronym? non acronym? 
}
Then, this paper defines compiler composition and formalises what kind of security properties does the composition of two compiler passes preserve, given what the individual passes preserve.
The paper then defines several secure compiler passes, where each is either preserving a different security property (e.g., temporal or spatial memory safety) or performing a security-preserving optimisation, (e.g., applying constant-folding or dead-code elimination).
\MP{
	acro? non acro?
}
Finally, this paper shows how to compose these secure compiler passes into a multi-pass compilation chain which is proven to provide end-to-end preservation of general memory safety and cryptographic constant time.
Crucially, this end-to-end security preservation is obtained by having each compilation pass (i) preserve a sub-part of the overall security property individually and then (ii) compose each preserved security property as dictated by the framework of this paper.
The results showcase how the framework allows the kind of formal security reasoning that compiler writers already want (and do), obtaining precise, compositional security reasoning while providing minimal (and modular) proof effort.

@@

  \item %
        %\Cref{sec:compcomp} studies different forms of secure compiler compositions.
        This paper takes the secure compilation framework of~\citep{abate2019jour} and extends it to reason about the security of all different known forms of compiler composition (\Cref{sec:compcomp}).
        %Concretely, it investigates upper and lower compositions, as well as the sequential composition.
        For this, this paper studies compilers with multiple inputs or multiple outputs as well as the sequential composition of compilers, as typically done in existing compilation chains.
        %
        This paper proves that starting from two compilers that preserve two (possibly distinct) properties, their composition preserves the intersection of those properties.
        For example, a compiler that is secure with respect to \gls{tms} can be composed with a compiler that is secure with respect to \gls{sms}, resulting in a compiler that is secure with respect to \gls{ms}.
%       \MPin{
%      		find way to connect to previous contrib
%       }
        % 
        Finally, this paper provides an important corollary that demonstrates that the order of composition of sequential compiler passes is irrelevant for the resulting security.
        % 
        This corollary is crucial for reordering compiler optimisation passes as well as secure compilation passes and thus generating secure and efficient code.

  \item %
        This paper presents a case-study showcasing the conjunction of the previous contributions (\Cref{sec:casestud:defs,sec:casestud:rtp}).
        To this end, it presents a compilation chain consisting of several passes.
        The chain ultimately preserves a combination of \gls{ms} and \gls{cct} by means of composing the individual, secure passes concerning \gls{tms}, \gls{sms}, and a stricter variant of \gls{cct}, respectively.
        Furthermore, the chain includes two passes for \gls{ms} that are optimising: One performs a simple \gls{dce} and the other \gls{cf}.
        The formalisation of this case study showcases the power of the presented framework: Namely, the divide-and-conquer approach to software engineering is a viable strategy even for the development of secure compilers.
%        This demonstrates that the usage of the presented framework is both simpler and more expressive compared to one monolithic security proof of the whole compilation chain.
